---
title: "Code for Davies et al, 2025"
author: "Kimberley Davies & Nancy Briggs"
date: "`r Sys.Date()`"
output: html_document
---

This document contains the code for the analysis reported in Davies, et al, 2025.

Davies, K, Lappin, JM, Briggs, N, Isobel, S, Steel, Z. Does shame mediate the influence of trauma on psychosis? A systematic review and meta-analytic structural equation modelling approach, _Schizophrenia Research_, 275, 87-97,
https://doi.org/10.1016/j.schres.2024.12.008.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


options(warnPartialMatchAttr = FALSE, 
  warnPartialMatchDollar = FALSE,
  warnPartialMatchArgs = FALSE)
```


```{r, message=FALSE, warning=FALSE}
# install.packages("readxl")                
# install.packages("corpcor")                
# install.packages("metaSEM")  
# install.packages("OpenMx")
# install.packages("MASS")
# install.packages("semPlot")
# install.packages("tidyverse")
# install.packages("meta")
# install.packages("metafor")
# install.packages("esc")
# install.packages("clubSandwich")

library(readxl)
library(corpcor)
library(metaSEM)
library(OpenMx)
library(MASS)
library(semPlot)
library(tidyverse)
library(dmetar)
library(meta)
library(metafor)
library(esc)
library(clubSandwich)


```

# Import data

```{r}

# You may need to edit the file path to import the data.
MASEM_data_model1 <- read_excel("../PublicationFiles/MASEM_data_model1.xlsx")

```



# Examine heterogeneity 
## Psychosis*Shame
```{r}

MASEM_data_model1<- MASEM_data_model1 %>%
  group_by(Study, Label) %>% 
  mutate(id_no = row_number())

# random effect for Correlation within study

res_masem_PS<-rma.mv(r_z, r_z_var,
                     subset=(Label=="PS"),
                     random = ~ 1 | Study/id_no,
                     data=MASEM_data_model1)
#res_masem_PS


W <- diag(1/res_masem_PS$vi)
X <- model.matrix(res_masem_PS)
P <- W - W %*% X %*% solve(t(X) %*% W %*% X) %*% t(X) %*% W

# Between-cluster heterogeneity
100 * sum(res_masem_PS$sigma2) / (sum(res_masem_PS$sigma2) + (res_masem_PS$k-res_masem_PS$p)/sum(diag(P)))

# Within-cluster heterogeneity
100 * res_masem_PS$sigma2 / (sum(res_masem_PS$sigma2) + (res_masem_PS$k-res_masem_PS$p)/sum(diag(P)))

```

## Psychosis*Childhood Trauma

```{r}

res_masem_PCT<- rma.mv(r_z, r_z_var,
                       subset=(Label=="PCT"),
                       random = ~ 1 | Study/id_no,
                       data=MASEM_data_model1)
#res_masem_PCT

W <- diag(1/res_masem_PCT$vi)
X <- model.matrix(res_masem_PCT)
P <- W - W %*% X %*% solve(t(X) %*% W %*% X) %*% t(X) %*% W

# Between-cluster heterogeneity
100 * sum(res_masem_PCT$sigma2) / (sum(res_masem_PCT$sigma2) + (res_masem_PCT$k-res_masem_PCT$p)/sum(diag(P)))

# Within-cluster heterogeneity
100 * res_masem_PCT$sigma2 / (sum(res_masem_PCT$sigma2) + 
                                   (res_masem_PCT$k-res_masem_PCT$p)/sum(diag(P)))
```

## Psychosis*Lifetime PTE

```{r}

MASEM_data_model1<- MASEM_data_model1 %>%
  group_by(Study, Label) %>% 
  mutate(id_no = row_number())

# random effect for Correlation within study
res_masem_PLT<-rma.mv(r_z, r_z_var,
                     subset=(Label=="PLT"),
                     random = ~ 1 | Study/id_no,
                     data=MASEM_data_model1)
#res_masem_PLT

W <- diag(1/res_masem_PLT$vi)
X <- model.matrix(res_masem_PLT)
P <- W - W %*% X %*% solve(t(X) %*% W %*% X) %*% t(X) %*% W

# Between-cluster heterogeneity
100 * sum(res_masem_PLT$sigma2) / (sum(res_masem_PLT$sigma2) + (res_masem_PLT$k-res_masem_PLT$p)/sum(diag(P)))

# Within-cluster heterogeneity
100 * res_masem_PLT$sigma2 / (sum(res_masem_PLT$sigma2) + (res_masem_PLT$k-res_masem_PLT$p)/sum(diag(P)))

```

## Shame*Lifetime PTE

```{r}

MASEM_data_model1<- MASEM_data_model1 %>%
  group_by(Study, Label) %>% 
  mutate(id_no = row_number())

# random effect for Correlation within study
res_masem_SLT<- rma.mv(r_z, r_z_var,
                       subset=(Label=="SLT"),
                       random = ~ 1 | Study/id_no,
                       data=MASEM_data_model1)
#res_masem_SLT

W <- diag(1/res_masem_SLT$vi)
X <- model.matrix(res_masem_SLT)
P <- W - W %*% X %*% solve(t(X) %*% W %*% X) %*% t(X) %*% W

# Between-cluster heterogeneity
100 * sum(res_masem_SLT$sigma2) / (sum(res_masem_SLT$sigma2) + (res_masem_SLT$k-res_masem_SLT$p)/sum(diag(P)))

# Within-cluster heterogeneity
100 * res_masem_SLT$sigma2 / (sum(res_masem_SLT$sigma2) + (res_masem_SLT$k-res_masem_SLT$p)/sum(diag(P)))
```

## Shame*Childhood PTE

```{r}

MASEM_data_model1<- MASEM_data_model1 %>%
  group_by(Study, Label) %>% 
  mutate(id_no = row_number())

# random effect for Correlation within study
res_masem_SCT<-rma.mv(r_z, r_z_var,
                      subset=(Label=="SCT"),
                      random = ~ 1 | Study/id_no,
                      data=MASEM_data_model1)
#res_masem_SCT

W <- diag(1/res_masem_SCT$vi)
X <- model.matrix(res_masem_SCT)
P <- W - W %*% X %*% solve(t(X) %*% W %*% X) %*% t(X) %*% W

# Between-cluster heterogeneity
100 * sum(res_masem_SCT$sigma2) / (sum(res_masem_SCT$sigma2) + (res_masem_SCT$k-res_masem_SCT$p)/sum(diag(P)))

# Within-cluster heterogeneity
100 * res_masem_SCT$sigma2 / (sum(res_masem_SCT$sigma2) + (res_masem_SCT$k-res_masem_SCT$p)/sum(diag(P)))
```



# Models

This code is adapted from Wilson et al. (2016)

Wilson, S. J., Polanin, J. R., & Lipsey, M. W. (2016). Fitting meta-analytic structural equation models with complex datasets. Research synthesis methods, 7(2), 121–139. https://doi.org/10.1002/jrsm.1199.

## Convert n to inverse sample size weighting

```{r}
MASEM_data_model1$inv_n <- 1/MASEM_data_model1$n
```

## Create multilevel model

The code calculates the correlation coefficients for the effects, allowing for multiple effects from a single study.


Table 3. Pooled correlation matrix for childhood and lifetime PTEs.

```{r}

 # constant sampling correlation assumption
rho <- 0.6

# constant sampling correlation working model
V <- with(MASEM_data_model1, 
          impute_covariance_matrix(vi = inv_n,
                                   cluster = Study,
                                   r = rho))

# Multilevel Random Effects Model
Step1 <-rma.mv (yi=ES, V=V,
                data=MASEM_data_model1,
                random = list(~1|Study, ~1|ES_ID),
                method = "ML", mods=~factor(Cell)-1)

summary (Step1)
```

## Extract variance-covariance matrix 

```{r}
Acov_model1 <-Step1$vb
```

## Create correlation matrix



Uses the calculated correlation coefficients to create a matrix.

```{r}
Matrix_model1 <-vec2sm (vec = c( 1, Step1$b[1], Step1$b[2], Step1$b[3],
                                 1, Step1$b[4], Step1$b[5],
                                 1, Step1$b[6],
                                 1),
                         diag=TRUE)
colnames(Matrix_model1) <- c("Psychosis", "Shame", "Lifetime Trauma", "Childhood Trauma")
rownames(Matrix_model1) <- c("Psychosis", "Shame", "Lifetime Trauma", "Childhood Trauma")

print(Matrix_model1)
```

# Enter sample size - this is total sample of all studies
```{r}
n<-3896
```




## Model 1 in paper

### A-Matrix

Defines paths to be estimated from Shame to Psychosis, Lifetime Trauma to Shame and Childhood Trauma to Shame.

```{r}
A_model1 <- create.mxMatrix(c(0, 0, 0, 0,
                              "0.2*Shame_Psy", 0, 0, 0,
                              0, "0.2*LT_Shame", 0, 0,
                              0, "0.2*CT_Shame", 0, 0),
                              ncol=4, nrow=4)

dimnames(A_model1)[[1]] <- dimnames(A_model1)[[2]] <- c("Psychosis", "Shame", "Lifetime Trauma", "Childhood Trauma")

A_model1 

```


### S-Matrix

Defines variances to be estimated, and variances fixed to 1.0

```{r}

S_model1<- create.mxMatrix(c("0.1*VarPsychosis", 0, 0, 0,
                             0, "0.1*VarShame", 0, 0,
                             0, 0, 1, 0,
                             0, 0, 0, 1),
                             ncol=4, nrow=4)

dimnames(S_model1)[[1]] <- dimnames(S_model1)[[2]] <- c("Psychosis", "Shame", "Lifetime Trauma", "Childhood Trauma")
S_model1

```




### Run metaSEM model


```{r}
model1.metaSEM<-wls(Cov=Matrix_model1, 
                    aCov=Acov_model1, 
                    n=n,
                    Amatrix=A_model1,
                    Smatrix=S_model1, 
                    cor.analysis=TRUE, 
                    diag.constraints=TRUE,
                    intervals.type = "LB",
                    mx.algebras = list(
                         TotindEffect = mxAlgebra(LT_Shame*Shame_Psy + CT_Shame*Shame_Psy,
                                       name="TotindEffect" ),
                         LPTEindirectEffect = mxAlgebra(LT_Shame*Shame_Psy , name="LPTEindirectEffect"),
                         CPTEindirectEffect = mxAlgebra(CT_Shame*Shame_Psy , name="CPTEindirectEffect"),
                         CompEffect = mxAlgebra(LT_Shame*Shame_Psy - CT_Shame*Shame_Psy,
                                                        name="CompEffect")))

summary(model1.metaSEM)

```



### Path Diagram
```{r}
sem.plot.model1 <- meta2semPlot(model1.metaSEM)


# Create Labels (left to right, bottom to top)
labels <- c("Psychosis", "Shame", "Lifetime Trauma", "Childhood Trauma")

# Plot
semPaths(sem.plot.model1, 
         whatLabels = "est", 
         edge.color = "black", 
         layout="tree", 
         rotation=2,
         nodeLabels = labels)

```


## Model 2 in paper

### A-matrix

```{r}
A_model2 <- create.mxMatrix(c(0, 0, 0, 0,
                              0, 0, 0, 0,
                              "0.2*LT_Psy", "0.2*LT_Shame", 0, 0,
                              "0.2*CT_Psy", "0.2*CT_Shame", 0, 0),
                              ncol=4, nrow=4)

dimnames(A_model2)[[1]] <- dimnames(A_model2)[[2]] <- c("Psychosis", "Shame", "Lifetime Trauma", "Childhood Trauma")

A_model2

```

### S-matrix


```{r}

S_model2 <- create.mxMatrix(c("0.1*VarPsychosis", 0, 0, 0,
                              0, "0.1*VarShame", 0, 0,
                              0, 0, 1, 0,
                              0, 0, 0, 1),
                     ncol=4, nrow=4)

dimnames(S_model2)[[1]] <- dimnames(S_model2)[[2]] <- c("Psychosis", "Shame", "Lifetime Trauma", "Childhood Trauma")
S_model2


```

### Run metaSEM Model 


```{r}
model2.metaSEM<-wls(Cov=Matrix_model1, 
                    aCov=Acov_model1, 
                    n=n,
                    Amatrix=A_model2,  
                    Smatrix=S_model2, 
                    cor.analysis=TRUE, 
                    diag.constraints=TRUE, 
                    intervals.type = "LB") 
 
summary(model2.metaSEM)
summary(model2.metaSEM, standardized=TRUE)

```

### Path Diagram



```{r}
sem.plot.model2 <- meta2semPlot(model2.metaSEM)


# Create Labels (left to right, bottom to top)
labels <- c("Psychosis", "Shame", "Lifetime Trauma", "Childhood Trauma")

# Plot
semPaths(sem.plot.model2, 
         whatLabels = "est", 
         edge.color = "black", 
         layout="tree", 
         rotation=2,
         nodeLabels = labels)

```



## Model 3 

```{r}

 # constant sampling correlation assumption
rho <- 0.6

# constant sampling correlation working model
V <- with(MASEM_data_model1, 
          impute_covariance_matrix(vi = inv_n,
                                   cluster = Study,
                                   r = rho))

Step1_model3 <-rma.mv (yi=ES, V=V,
                data=MASEM_data_model1,
                random = list(~1|Study, ~1|ES_ID),
                method = "ML", mods=~factor(Cell)-1)




summary (Step1_model3)



```

### Extract variance-covariance matrix 

```{r}
Acov_model3 <-Step1_model3$vb

```

### Create correlation matrix 
```{r}
Matrix_model3 <-vec2sm (vec = c(1, Step1_model3$b[1], Step1_model3$b[2], Step1_model3$b[3],
                                 1, Step1_model3$b[4], Step1_model3$b[5],
                                 1, Step1_model3$b[6],
                                 1),
                         diag=TRUE)

colnames(Matrix_model3) <- c("Psychosis", "Shame", "Lifetime Trauma", "Childhood Trauma")
rownames(Matrix_model3) <- c("Psychosis", "Shame", "Lifetime Trauma", "Childhood Trauma")

print(Matrix_model3)

```

### Enter sample size

```{r}
n<-3896
```


### A-Matrix

```{r}
A_model3 <- create.mxMatrix(c(0, 0, 0, 0,
                              "0.2*Shame2Psy", 0, 0, 0,
                              "0.2*LT2Psy", "0.2*LT2Shame", 0, 0, 
                              "0.2*CT2Psy", "0.2*CT2Shame", 0, 0),
                              ncol=4, nrow=4)

dimnames(A_model3)[[1]] <- dimnames(A_model3)[[2]] <- c("Psychosis", "Shame", "Lifetime Trauma", "Childhood Trauma")
A_model3
```

### S-Matrix 
```{r}
S_model3<- create.mxMatrix(c("0.1*errPsychosis", 0, 0, 0,
                             0, "0.1*errShame", 0, 0,
                             0, 0, 1, 0,
                             0, 0, 0, 1),
                             ncol=4, nrow=4)

dimnames(S_model3)[[1]] <- dimnames(S_model3)[[2]] <- c("Psychosis", "Shame", "Lifetime Trauma", "Childhood Trauma")
S_model3

```



### Run metaSEM model 

```{r}
model3.metaSEM<-wls(Cov=Matrix_model3, 
                    aCov=Acov_model3, 
                    n=n, 
                    Amatrix=A_model3,
                    Smatrix=S_model3, 
                    cor.analysis=TRUE, 
                    diag.constraints=TRUE,
                    intervals.type = "LB",
                    mx.algebras = list( TotalindirectEffect = mxAlgebra(LT2Shame*Shame2Psy + CT2Shame*Shame2Psy, name="TotalindirectEffect" ),
                         LTindirectEffect = mxAlgebra(LT2Shame*Shame2Psy , name="LTindirectEffect" ) ,
                         CTindirectEffect = mxAlgebra(CT2Shame*Shame2Psy , name="CTindirectEffect" ) ,
                         CompindirectEffect = mxAlgebra(LT2Shame*Shame2Psy - CT2Shame*Shame2Psy,
                                       name="CompindirectEffect")))

summary(model3.metaSEM)
```



### Path Diagram

```{r}

sem.plot.model3 <- meta2semPlot(model3.metaSEM)


# Create Labels (left to right, bottom to top)
labels <- c("Psychosis", "Shame", "Lifetime Trauma", "Childhood Trauma")

# Plot
semPaths(sem.plot.model3, 
         whatLabels = "est", 
         edge.color = "black", 
         layout="tree", 
         rotation=2,
         nodeLabels = labels)
```

# Publication Bias



## Multilevel Eggers test 

```{r}
# Egger's test (Eq 21, Nakagawa, 2021)
mod_ma <- rma.mv(r_z, r_z_var, mod = ~ sqrt(r_z_var), 
                 random = list(~1|Study, ~1|ES_ID), 
                 data = MASEM_data_model1)
summary(mod_ma)
# Because slope of the SE effect is p>0.05, there is no
# evidence of small study bias according to the multilevel test

# This is not needed becasue there is no relationship in the above model.
mod_mb <- rma.mv(r_z, r_z_var,  mod = ~ r_z_var, 
                 random = list(~1|Study, ~1|ES_ID), 
                 data = MASEM_data_model1)
summary(mod_mb)
```



## Test Risk of bias as a moderator 

```{r}
che.model.bias <- rma.mv(ES ~ 1 + relevel(factor(bias), ref="Poor"),
                    V = V,
                    random = ~ 1 | Author/ES_ID,
                    data = MASEM_data_model1,
                    sparse = TRUE)


conf_int(che.model.bias, vcov = "CR2")
coef_test(che.model.bias, vcov = "CR2")
summary(che.model.bias)
```


## Funnel Plot

```{r, fig.height=8}
# Adapted from https://sakaluk.wordpress.com/2016/02/16/7-make-it-pretty-plots-for-meta-analysis/



### fit equal-effects model
res_masem_pb <- rma(r_z, r_z_var, data=MASEM_data_model1, measure="COR", method="EE")



# Store the meta-analytic estimate and its standard error
estimate = res_masem_pb$beta[1]
se = res_masem_pb$se[1]
max_se_val <- max(sqrt(MASEM_data_model1$r_z_var))
se.seq=seq(0, max_se_val, 0.001)

# Compute vectors of the lower-limit and upper limit values for
# the 95% CI region, using the range of SE generated in the 
# previous step, and the stored value of the meta-analytic estimate.

ll95 = estimate-(1.96*se.seq)
ul95 = estimate+(1.96*se.seq)

# 99% CI region too

ll99 = estimate-(3.29*se.seq)
ul99 = estimate+(3.29*se.seq)

# Do the same thing except now calculating the confidence interval
# for your meta-analytic estimate based on the stored value of the
# standard error

meanll95 = estimate-(1.96*se)
meanul95 = estimate+(1.96*se)

# Make dataframe
dfCI = data.frame(ll95, ul95, ll99, ul99, se.seq, estimate, meanll95, meanul95)

# create variable that is concatenation of Correlation and Author
MASEM_data_model1$Corr_auth<-paste(MASEM_data_model1$Label, ", ", MASEM_data_model1$Author)

 

# Create the funnel plot.
fp =
  ggplot(aes(x = sqrt(MASEM_data_model1$r_z_var), y = ES), data = MASEM_data_model1) +
  geom_point(size=2, aes(shape=Label, colour=Author) ) +
  xlab('Standard Error') + 
  ylab('Correlation Coefficient')+
#CI regions
  geom_line(aes(x = se.seq, y = ll95), linetype = 'dotted', data = dfCI) +
  geom_line(aes(x = se.seq, y = ul95), linetype = 'dotted', data = dfCI) +
# line for the estimate itself.
   geom_segment(aes(x = min(se.seq), y = estimate, 
                    xend = max(se.seq),
                    yend = estimate), 
                linetype='dotted', data=dfCI) +
  scale_x_reverse()+   
#Specify the range and interval for the tick-marks of the y-axis (Zr);
  scale_y_continuous(breaks=seq(-1,2,0.2))+
  labs(shape="Correlation Type") +
  coord_flip() +   # Flip the x and y axes
  theme_classic( ) +
  theme (
    legend.text = element_text(size = 6),
    legend.title = element_text(size = 8),
    legend.spacing.y = unit(.5, "cm")
  )
fp 


```


```{r}
sessionInfo()
```
